---
title: "HW6"
author: "Richard Gao - PID:A16490010"
date: "2024-02-03"
output: pdf_document
---

```{r}
# install.packages("bio3d")
# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

```{r}
View(s1.b)
```


## This function will plot the bfactor values of a particular chain of amino 
## acids of a desired protein against the protein chain residue locations. You 
## will have to input the protein file, the chain of interest, and the specific 
## atom you are using as an electrostatic marker for the bfactor values 
## (usually "CA"). The function will output a line graph where the x-axis is 
## the residue location based on the chosen electrostatic marker atoms, and the 
## y-axis is the bfactor which reflects the flexibility of the residue at that 
## location/atom.
```{r}
# input a protein data bank coordinate file, a string variable of the desired 
# residue, and a string variable of the elety property
plotProtein <- function(proteinFile, chainName, eletyProperty) {
  
  # reads the protein data bank coordinate file and stores it as a list 
  # "proteinData" of class "pdb" containing the structural data of the protein
  proteinData <- read.pdb(proteinFile)
  
  # selects a particular subset of atoms from the "proteinData" pdb object and 
  # stores is as a smaller pdb object "proteinChain"
  proteinChain <- trim.pdb(proteinData, chain = chainName, elety=eletyProperty)
  
  # stores the Bfactor value of each amino acid in the desired chain as a 
  # vector "BfactorValues
  BfactorValues = proteinChain$atom$b
  
  # creates a line plot of the Bfactor (which is reflective of stability) of 
  # each amino acid residue in the desired protein chain
  plotb3(BfactorValues, sse=proteinChain, typ = "l", ylab = "Bfactor")
}
```

```{r}
plotProtein("4AKE", "A", "O")
```

```{r}
plot(s1.b)
```


