---
title: "Class18: Investigating Pertussis Resurgence"
author: "Richard Gao (PID: A16490010)"
format: pdf
---

First we will examine and explore Pertussis cas numbers in the US as tracked by the CDC:
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

We can use the datapastapackage to scrape this data from the website into R:

```{r}
cdc <- data.frame( year = c(1922L,1923L,1924L,1925L, 1926L,1927L,1928L,1929L,1930L,1931L,1932L,1933L,1934L,1935L,1936L,1937L,1938L,1939L,1940L,1941L,1942L,1943L,1944L,1945L,1946L,1947L,1948L,1949L,1950L,1951L,1952L,1953L,1954L,1955L,1956L,1957L,1958L,1959L,1960L,1961L,1962L,1963L,1964L,1965L,1966L,1967L,1968L,1969L,1970L,1971L,1972L,1973L,1974L,1975L,1976L,1977L,1978L,1979L,1980L,1981L,1982L,1983L,1984L,1985L,1986L,1987L,1988L,1989L,1990L,1991L,1992L,1993L,1994L,1995L,1996L,1997L,1998L,1999L,2000L,2001L,2002L,2003L,2004L,2005L,2006L,2007L,2008L,2009L,2010L,2011L,2012L,2013L,2014L,2015L,2016L,2017L,2018L,2019L,2020L,2021L),
  cases = c(107473,164191,165418,152003,202210,181411,161799,197371,166914,172559,215343,179135,265269,180518,147237,214652,227319,103188,183866,222202,191383,191890,109873,133792,109860,156517,74715,69479,120718,68687,45030,37129,60886,62786,31732,28295,32148,40005,14809,11468,17749,17135,13005,6799,7717,9718,4810,3285,4249,3036,3287,1759,2402,1738,1010,2177,2063,1623,1730,1248,1895,2463,2276,3589,4195,2823,3450,4157,4570,2719,4083,6586,4617,5137,7796,6564,7405,7298,7867,7580,9771,11647,25827,25616,15632,10454,13278,16858,27550,18719,48277,28639,32971,20762,17972,18975,15609,18617,6124,2116)
       )
```

```{r}
head(cdc)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)
```

```{r}
ggplot(cdc, aes(year, cases)) +
  geom_line() +
  geom_vline(xintercept=1947, col="blue") +
  geom_vline(xintercept=1992, col="red") +
  geom_vline(xintercept=2020, col="green") 
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

I notice that between the 1946 introduction of the wP vaccine and the 1996 switch to the aP vaccine, the number of pertussis cases dropped significantly and stayed low.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, the number of perstussis cases resurged. A possible explanation for this is that the immunity gained from the aP vaccine doesn't last very long compared to the original wP vaccine.


Access data from the CMI-PB project

This database (like many modern projects) uses an APi to resturn JSON format data. 

We will use the R package `jsonlite`.

```{r}
library(jsonlite)
subject <- read_json("http://cmi-pb.org/api/subject",
                     simplifyVector = TRUE)
head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
sum(subject$infancy_vac == "wP")
sum(subject$infancy_vac == "aP")
```
or just do:
```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
subject$year_of_birth
```

# Side-Note: Working with dates

We can use the lubridate package to ease the pain of doing math with dates.

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2002-11-19")
```


```{r}
time_length(today() - ymd("2002-11-19"), "years")
```

So what is the age of everyone on our dataset.

```{r}
subject$age <- time_length( today() - ymd(subject$year_of_birth), "years" )
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)
wPsubject <- filter(subject, infancy_vac == "wP")
head(wPsubject)
mean(wPsubject$age)

aPsubject <- filter(subject, infancy_vac == "aP")
mean(aPsubject$age)

t.test(wPsubject$age, aPsubject$age)
```
Average age of wP individuals is 36.58 yrs old and that of aP individuals is 26.28 yrs old. Yes they are significantly different with a p-value < 2.2e-16.

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

```{r}
ggplot(subject) +
  aes(age, fill = infancy_vac) +
  geom_histogram()
```
> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2)
```

Yes, those who received the wP vaccine seem to be significantly older than those who received the aP vaccine.

## Get more data from CMI-PB

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```

We need to **join** these two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the `dplyr` join functions to do this:

> Q10. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer",
                      simplifyVector = T)

head(ab_titer)
```

One more `inner_join()` to add all our metadata in `meta` on to our `ab_data` table:

> Q11. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
dim(abdata)
```

Our first exploratory plot:

> Q12. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$antigen)
```

> Q13. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
We have less rows for more recent years.

## Examine IgG Ab titer levels

Now using our joined/merged/linked abdata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q14. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q15. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT, PRN, FIM2/3, and FHA. These were included in the vaccine.


```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

> Why are certain antigens and not others very variable in their detected levels here?

The vaccine only contains certain antigens.

> Can you facet or even just color by infancy_vac? Is there some difference?

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(infancy_vac), nrow=2)
```
There are potentially some differences here but in general it is hard to tell with this whole dataset overview...
```{r}
table(abdata$dataset)
```

> Q16. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q17. What do you notice about these two antigens time courses and the PT data in particular?

PT increases for a while then has a sharp drop off while FIM 2/3 steadily increases over time and only slightly decreases.


Let's focus in on just the 2021_dataset.

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
table(abdata.21$dataset)
```

Focus on PT antigen for IgG levels
```{r}
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")
```

Plot of
```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,) +
    geom_point()
```

```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
    labs(title="2021 Dataset IgG PT", x = "Days After Boost", y = "MFI")
```

> Q18. Do you see any clear difference in aP vs. wP responses?

wP responses show a greater peak at 14-days post-vaccination.








