---
title: "Class16 - R Graph"
author: "Richard Gao (PID: A16490010)"
format: pdf
---

```{r}
data <- read.table("~/Desktop/class16/mm-second.x.zebrafish.tsv")
colnames(data) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
```

Make a histogram of the $bitscore values. You may want to set the optional breaks to be a larger number (e.g. breaks=30).
```{r}
hist(data$bitscore, breaks=30)
```
> What do you notice here? Note that larger bitscores are better.

I notice that the histogram is bimodal and right-skewed, there are not many larger bitscores which makes sense since they are "better" so will probably be rarer.

```{r}
library(ggplot2)
ggplot(data, aes(pident, bitscore)) + geom_point(alpha=0.1) 
```
> Is there a straightforward relationship between percent identity ($pident) and bitscore ($bitscore) for the alignments we generated?

No there seems to be a slight positive correlation but there must be another factor involved (length of alignment).


```{r}
ggplot(data, aes((data$pident * (data$qend - data$qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth()
```


